A putative causal map of natural variation across the human phenome

### 2017-08-03 docs/manuscript.tex
updated

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Learning pleiotropic signatures to construct a causal map of the human phenome},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\usepackage{amsmath}

@@

\setlength{\droptitle}{-2em}
  \title{Learning pleiotropic signatures to construct a causal map of the human
phenome}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{02 August 2017}

@@

A major application for genome-wide association studies (GWAS) has been
the emerging field of causal inference using Mendelian randomization
(MR), where the causal effect between a pair of traits can be estimated
using only summary level data. MR depends on SNPs exhibiting vertical
pleiotropy, where the SNP influences an outcome phenotype only through
an exposure phenotype. Issues arise when this assumption is violated due
to SNPs exhibiting horizontal pleiotropy, and many methods have been
developed in an attempt to address this, each assuming a different model
of horizontal pleiotropy. Here we show that across a range of horizontal
pleiotropy models instrument selection will be increasingly liable to
selecting invalid instruments as GWAS sample sizes continue to grow. We
have integrated these methods into a mixture-of-experts machine learning
approach (MR-MoE 1.0) that predicts the most appropriate model to use
for any specific causal analysis, improving on both power and false
discovery rates. Using the approach, we systematically estimated the
causal effects amongst 2407 phenotypes, generating a working draft of
the causal map of the human phenome, comprising the evaluation of
causality between over a million pairs of traits. Almost 90\% of causal
estimates indicated some level of horizontal pleiotropy. The causal
estimates are organised into a publicly available graph database
(\url{http://eve.mrbase.org}).

@@

Problems with obtaining unbiased causal effects can arise, however, if
the genetic instruments exhibit horizontal pleiotropy (HP), where they
influence the outcome through a pathway other than the exposure (2). The
extent of this phenomenon is not to be understated, and many methods
have been developed that attempt to reliably obtain unbiased causal
estimates under specific models of HP (4--7,9,13). It is considered best
practice to report estimates from a range of methods as a sensitivity
analysis when presenting causal estimates so as to assess the potential
for different horizontal pleiotropy models and other violations of MR
assumptions. However it is also desirable to consolidate across methods
under some circumstances. First, it could be of interest to screen links
between thousands of traits for being potential causal associations, in
which case a critical evaluation of each causal estimate of interest may
not be possible or convenient. Second, though the simplest method, based
on inverse-variance weighted (IVW) meta analysis (11), is most
statistically powerful under no HP, it can have high false positive or
low true positive rates in the presence of HP compared to other methods.
Pleiotropy has been hypothesised to be universal (14), though the degree
and nature (e.g.~horizontal versus vertical) to which this may be the
case is contested (15,16), hence defaulting to the IVW method in the
first instance and using other methods as sensitivity analyses may not
be appropriate. Third, if different methods disagree it is not possible
to know which is correct because the true nature of HP exhibited by the
instruments is not known. Fourth, the available methods do not cover all
possible models of HP, and therefore an automated method for instrument
selection may be necessary.

In this paper we introduce two innovations towards improving the
reliability of MR estimates. First, we introduce an approach to discard
genetic variants that are likely to be invalid. Second, we hypothesised
that characteristics of the summary data could indicate which method
would be most reliable, and we introduce new machine learning approaches
that attempt to automate both instrument and method selection. Using
curated GWAS summary data for thousands of phenotypes (12), we use these
new methods to construct a graph of millions of causal estimates.
Motivated by the recent avalanche (17) of 2SMR publications, with
similar (18,19) or contradicting (20,21) conclusions despite using the
same data (17), we developed a graph database to represent these
estimates in a consistent manner. We consider this to be a `working
draft' of the causal map of the human phenome, but raise caution
throughout that its interpretation is far from straightforward, and that
future corrections and refinements will inevitably follow as data grows
and 2SMR methods evolve.

\subsubsection{Box 1: Models of
pleiotropy}\label{box-1-models-of-pleiotropy}

In this paper we separate pleiotropy into two forms for simplicity -
vertical and horizontal. Vertical pleiotropy has also been known as
mediated pleiotropy (22), because the genetic variant influences one
trait, which in turn influences a second trait. It has also been termed
Type II pleiotropy (15) and secondary pleiotropy (23).

By contrast, horizontal pleiotropy can manifest in a myriad of different
ways, but the guiding principle is that the genetic variant influences
two different phenotypes through independent pathways. Horizontal
pleiotropy has also been termed biological pleiotropy (22), Type I
pleiotropy (15), developmental pleiotropy or selectional pleiotropy
(24). From a MR statistical perspective, one can be more specific and
state that horizontal pleiotropy occurs where a SNP influences an
outcome trait through some pathway other than the exposure trait.

For illustration, Hu et al (2016) (Figure 3) enumerate the biological
mechanisms that can lead to pleiotropy: a single variant influences
multiple genes; a single gene generates multiple gene products; a single
gene product has multiple cellular functions; or a single cellular
function influences multiple phenotypes (25). Whether we consider each
of these mechanisms to be vertical pleiotropy (advantageous for MR) or
horizontal pleiotropy (problematic for MR) depends on which phenotypes
are being interrogated. For example, consider a scenario where a SNP
influences the expression levels of two genes, and the first gene
influences trait A while the second gene influences trait B. Then:

\begin{itemize}
\tightlist
\item
  The genetic variant is exhibiting a vertical pleiotropic effect on
  trait A through the first gene expression level. Here, all other
  assumptions being met, the genetic variant would be a valid instrument
  if used to estimate the causal effect of the gene expression level on
  trait A. However,
\item
  the SNP is exhibiting a horizontal pleiotropic effect on trait A and
  trait B through the two gene expression levels. Here the genetic
  variant would be invalid if used to estimate the causal effect of
  trait A on trait B.
\end{itemize}

@@

The use of summary data in two-sample MR is described in detail
elsewhere (12,26). A brief outline of the procedure is as follows.
First, genetic instruments for the exposure trait need to be identified
- those SNPs with \(p < 10^{-8}\) are retained in order to ensure that
the first assumption of MR (that the instrument associates with the
exposure) is generally satisfied. We collect their effect sizes,
standard errors and effect alleles for the association with the exposure
trait. These can be obtained from published GWAS summary data, either
from from curated lists of established GWAS associations (10) or study
websites comprising complete summary data (all SNP effects across the
genome regardless of strength of association). Next, the effects of
those SNPs on the outcome need to be obtained, typically necessitating
access to complete summary data because these SNPs will not usually
reach genome-wide significance for the outcome trait (and are therefore
usually absent in curated catalogues).

@@

The Rucker framework (27), adapted to MR (6) uses estimates of
heterogeneity in the IVW and Egger frameworks to navigate between these
nested models. A jackknife approach (random selection with replacement
of the complete set of instruments) can be used to obtain a sampling
distribution for the model estimate amongst these four variations. Using
1000 rounds of jackknife estimates, we can obtain a final estimate using
the mean or the median of the distribution. We only use the jackknife
approach for MR analyses where there are 15 or more instruments, in
order to avoid saturating the possible number of instrument
combinations.

@@

\textbf{Median-based methods:} An alternative approach is to take the
median effect of all available instruments (5,28). This has the
advantage that only half the instruments need to be valid, and the
estimate will remain unbiased. The weighted median estimate develops the
approach further to allow stronger instruments to contribute more
towards the estimate can be obtained by weighting the contribution of
each instrument by the inverse of its variance. The penalised weighted
median estimator introduces a further weight to the instruments,
penalising any instrument that contributes substantially towards the
heterogeneity statistic. Together, this provides three median based
estimators. Other estimation strategies not considered here, such as
LASSO regression, have also been developed for the situation where at
least half of the instruments are valid (29).

@@

The simplest approach to selecting instruments for performing MR is to
take SNPs that have been declared significant in the published GWAS for
the exposure. This typically involves obtaining SNPs that surpass
\(p < 5 \times 10^{-8}\), using clumping to obtain independent SNPs, and
then replicating in an independent sample. These results are often
recorded in public GWAS catalogs. Alternatively the clumping procedure
can be performed using complete summary data in MR-Base (12). Complete
summary data refers to association results for all SNPs used in a GWAS
i.e.~not only those passing statistical thresholds for significance. We
call this the ``tophits'' strategy.

@@

because the \(cor(g_{i}, B)^2 = cor(A, B)^{2} cor(g_{i}, A)^{2}\). This
simple inequality will not hold in some cases, for example
\(\rho_{x, x_o} < \rho_{x,y}\rho_{y,y_o}\) where \(\rho_{x, x_o}\) and
\(\rho_{y, y_o}\) are the precision of the measurements of the \(x\) and
\(y\). Some combinations of confounding effects can also distort the
\(\rho_{g,x}\) and \(\rho_{g,y}\) parameters, as has been discussed in
detail previously (8). However, we use it here as a computationally
inexpensive and approximate method to identify variants that are likely
to be invalid. Steiger's Z-test of correlated correlations (30) can be
used to formally test the extent to which the two correlations are
statistically different.

Here we adapt this approach to automatically filter SNPs that are liable
to be invalid (Figure 1a). Other methods have been developed to identify
invalid instruments for the purposes of exclusion from MR analyses
(9,13,31), based on the notion that outlying instruments are more likely
to be a source of HP. A potential drawback of this approach is that the
outlier SNPs in a summary-set might be the only reliable ones. This
could arise, for example, for the \emph{CRP} variant instrumenting
C-reactive protein levels (20), or the \emph{SLC2A9} variant influencing
serum urate levels (32).

@@

We consider the 14 MR methods described above, for which instruments can
be supplied using two instrument selection strategies, leading to 28
methods in total (Supplementary table 2). Each method assumes or
performs best for a different model of pleiotropy. Our objective is to
select the method most likely to be correct for a specific MR analysis
by predicting the model of pleiotropy, and relating that prediction to
the most appropriate model. To achieve this we use a ``mixture of
experts'' machine learning approach (33), where each MR method is
considered to be an `expert', taking a summary-set as its input. An
overview of the approach is shown in Figure 2.

@@

It should be noted that in the original hierarchical mixture of experts
approach of (33) the gating function and the experts share the same
input space and are trained simultaneously using
expectation-maximisation. In our approach the gating function is defined
over a separate input space consisting of the parameter space of the
experts, and is trained separately in a supervised setting using
simulated data. As such our MR-MoE approach fits the framework of
meta-learning (34--36) where machine learning methods are applied at the
`meta-level' to analyse the results of machine learning experiments at
the `base-level', in order to be able to understand and model the
capabilities of each expert and recommend which expert should be applied
to a given problem. From this perspective the 28 MR methods are the
base-models, the gating function is the meta-model, and its
meta-features are the parameters of the summary-sets.

@@

We applied our analysis using summary data for 2407 phenotypes,
including 149 complex traits and diseases, 575 metabolites (37,38) and
1683 plasma protein levels (39). For the protein levels only GWAS
significant hits were available, so they could only be evaluated as
exposure phenotypes. The complex traits and diseases and metabolite
levels had complete summary data available obtained via MR-Base, and
thus they could be evaluated as both exposures (if they had significant
instruments) and outcomes. Together, we evaluated 715681 relationships.
The majority of these associations could only be evaluated using fewer
than 5 SNPs, and so the Wald ratio or IVW fixed effects methods were
used, but for 61029 associations the MR-MoE approach was applied.

@@

As an example of the database we performed a look up of associations
between all traits and LDL cholesterol (as measured in the GLGC
consortium (40)), where the MoE result had a false discovery rate of
0.05 using the MoE approach. This returned 287 putative associations,
amongst which 111 involved LDL cholesterol influencing other traits and
176 involved other traits influencing LDL cholesterol. A large
proportion of these traits were metabolites, which are dominated by
lipid fractions. Filtering to exclude the metabolomic studies ((37,38))
returned 27 associations with 23 traits (Figure 4a). There was a strong
association with coronary heart disease (0.41 log(OR) per SD, 0.31-0.52,
\(p=2.7\times10^{-12}\)). Here the IVW random effects method was chosen
after Steiger filtering, indicating the presence of HP being present
amongst the instruments.

We also performed a look up of associations that causally influence
years of schooling (41), or that years of schooling itself causally
influences. Using a false-discovery rate of 0.05, 45 traits were
returned as having some direction of causality with years of schooling
(Figure 4b). Several of these putative relationships appear to be
plausible, for example the influence of years of schooling on lower risk
of Alzheimer's disease, which has recently been reported (42). However,
it is clear that there are serious limitations in relying on MR as a
panacea for causal inference. For example, bi-directional causal
relationships with college completion exist because of the similarity in
trait definitions; and there are several traits (e.g.~childhood
intelligence and birth weight) which appear to be causally downstream of
years of schooling despite this being temporally impossible. Though
explanations could be conjured for this association, for example
parents' schooling influences childhood intelligence, or the genetic
instruments are shared across the two traits due to a third unmeasured
trait, an intuitive interpretation of causality cannot be applied
because the genetic instruments used proxy the liability to achieve some
level of educational attainment, and not necessarily the experience of
having attained a particular level of edication.

@@

The trend of increasing GWAS sample sizes continues, but while the
opportunity that this affords MR to be furnished with more instruments
is typically welcomed, here we have shown that it is invalid
instruments, exhibiting horizontal pleiotropy or reverse causation, that
are more likely to be identified than valid ones. Strategising how MR is
to be used in practice, therefore, must consider horizontal pleiotropy
to be the rule rather than the exception. Following a recent flurry in
development of MR methodology, we have devised a machine learning
approach, MR-MoE, that seeks to predict the performance of each MR
method, selecting the one which is most likely to maximise power whilst
minimising FDR for a specific summary-set. The mixture of experts
approach that we present here is trained using random forest decision
trees applied to extensive and diverse simulations, and we demonstrate
that it makes substantial performance improvements over any other single
method in the presence or absence of extensive horizontal pleiotropy. As
such, our method contributes to the field of meta-learning but also to
the rapidly growing field of automatic model selection and
hyper-parameter optimization in machine learning (43).

Crucially however, we observed in our simulations that none of the
available methods were conservative and all, including MR-MoE, had FDR
above 0.2. Instrumental variables in MR are typically chosen blind, with
GWAS significance being the only criterion. We illustrated showed how
confounders and reverse causal associations can easily lead to invalid
instrument selection in MR and that most randomly generated patterns of
horizontal pleiotropy cannot be adequately accounted for by any single
MR method.

We applied the method to the curated set of GWAS summary data present in
MR-Base (12), generating several million MR estimates. MR-MoE selected a
method that indicated some pattern of horizontal pleiotropy for almost
90\% of cases, reinforcing the notion that it is the rule rather than
the exception. Based on our simulations we expect that those
associations that we reported to be `significant' are liable to high
type 1 error rates.

One advantage to generating a method that consolidates across many
methods is that it can prevent cherry picking the most appealing method
based on their results, but in general we strongly advocate the use of
reporting multiple MR methods to appraise any specific the impact of HP
on a causal estimate. In addition to HP, there are many other
limitations that prevent MR from being a panacea for causal inference.
Many of the associations are biologically impossible, for example where
early life phenotypes appear to be influenced by later stage phenotypes.
Such associations can be explained statistically because the instruments
proxy the liability for that particular phenotype, not that phenotype's
observed manifestation. Though these associations can be informative,
their interpretations as causal relationships are far from clear.
Similarly, often disease traits appear to causally relate to other
phenotypes, but GWAS of disease is typically performed on the liability
scale, hence the causal estimate reflects not the presence or absence of
disease, but the underlying risk of disease. Again, interpretation of
such associations can be problematic due to the instruments proxying
disease liability and not diseae outcomes.

>>>>

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={A putative causal map of natural variation across the human phenome},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\setlength{\droptitle}{-2em}
  \title{A putative causal map of natural variation across the human phenome}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{03 August 2017}

\usepackage{amsmath}

@@

A major application for genome-wide association studies (GWAS) has been
the emerging field of causal inference using Mendelian randomization
(MR), where the causal effect between a pair of traits can be estimated
using only summary level data. MR depends on SNPs exhibiting vertical
pleiotropy, where the SNP influences an outcome phenotype only through
an exposure phenotype. Issues arise when this assumption is violated due
to SNPs exhibiting horizontal pleiotropy. We demonstrate that across a
range of pleiotropy models, instrument selection will be increasingly
liable to selecting invalid instruments as GWAS sample sizes continue to
grow. Methods have been developed in an attempt to protect MR from
different patterns of horizontal pleiotropy, and here we have designed a
mixture-of-experts machine learning framework (MR-MoE 1.0) that predicts
the most appropriate model to use for any specific causal analysis,
improving on both power and false discovery rates. Using the approach,
we systematically estimated the causal effects amongst 2407 phenotypes,
comprising the evaluation of causality between over a million pairs of
traits. Almost 90\% of causal estimates indicated some level of
horizontal pleiotropy. The causal estimates are organised into a
publicly available graph database (\url{http://eve.mrbase.org}), and we
use it here to highlight the numerous challenges that remain in
automated causal inference.

@@

Problems with obtaining unbiased causal effects can arise, however, if
the genetic instruments exhibit horizontal pleiotropy (HP), where they
influence the outcome through a pathway other than the exposure (Box 1).
The extent of this phenomenon is not to be understated, and many methods
have been developed that attempt to reliably obtain unbiased causal
estimates under specific models of HP (4--7,9,13). It is considered best
practice to triangulate estimates from a range of MR methods (and other
experimental designs) when presenting causal estimates so as to assess
the potential for different horizontal pleiotropy models and other
violations of MR assumptions (14). However there are circumstances in
which it is desirable to consolidate across many methods to obtain a
single MR estimate. First, it could be of interest to screen links
between thousands of traits for being potential causal associations, in
which case a critical evaluation of each causal estimate of interest may
not be possible or convenient. Second, though the simplest method, based
on inverse-variance weighted (IVW) meta analysis (11), is most
statistically powerful under no HP, it can have high false positive or
low true positive rates in the presence of HP compared to other methods.
Pleiotropy has been hypothesised to be universal (15), though the degree
and nature (e.g.~horizontal versus vertical, see Box 1) to which this
may be the case is contested (16,17), hence defaulting to the IVW method
in the first instance and using other methods as sensitivity analyses
may not be appropriate. Third, if different methods disagree it is not
possible to know which is correct because the true nature of HP
exhibited by the instruments is not known. Fourth, the available methods
do not cover all possible models of HP, and therefore an automated
method for instrument selection may be necessary.

In this paper we introduce two innovations towards improving the
reliability of MR estimates. First, we present an approach to discard
genetic variants that are likely to be invalid. Second, we hypothesised
that characteristics of the summary data could indicate which method
would be most reliable, and we introduce new machine learning approaches
that attempt to automate both instrument and method selection. Using
curated GWAS summary data for thousands of phenotypes (12), we use these
new methods to construct a graph of millions of causal estimates.
Motivated by the recent avalanche (18) of 2SMR publications, with
similar (19,20) or contradicting (21,22) conclusions despite using the
same data (18), we developed a graph database to represent these
estimates in a consistent manner. We consider this to be a `working
draft' of the causal map of the human phenome, but raise caution
throughout that its interpretation is far from straightforward, and that
future corrections and refinements will inevitably follow as data grows
and 2SMR methods evolve.

\subsubsection{Box 1: Pleiotropic
mechanisms}\label{box-1-pleiotropic-mechanisms}

Pleiotropy is the phenomenon whereby a single genetic variant influences
multiple phenotypes. The mechanisms behind it can arise through many
different mechanisms, for example a single variant influences multiple
genes; a single gene generates multiple gene products; a single gene
product has multiple cellular functions; or a single cellular function
influences multiple phenotypes (23). How we classify the pleiotropic
mechanism depends on the application, and in this paper we discuss
pleiotropy in two forms - vertical and horizontal. Vertical pleiotropy
has also been known as mediated pleiotropy (24), because the genetic
variant influences one trait, which in turn influences a second trait.
It has also been termed, in some contexts, Type II pleiotropy (16) and
secondary pleiotropy (25). It is this form of pleiotropy that MR assumes
to make causal inference between two traits.

By contrast, horizontal pleiotropy can manifest in a myriad of different
ways, but the basic principle is that the genetic variant influences two
different phenotypes through independent pathways. Depending on context,
horizontal pleiotropy can also be termed biological pleiotropy (24),
Type I pleiotropy (16), developmental pleiotropy or selectional
pleiotropy (26). From the perspective of MR, horizontal pleiotropy
arises when a SNP influences an outcome trait through some pathway other
than the exposure trait.

A genetic variant can exhibit both horizontal and vertical pleiotropy in
different MR analyses. For example, consider a scenario where a SNP
influences the expression levels of two genes (X and Y), and gene X
influences trait A while gene Y influences trait B. Then:

\begin{itemize}
\tightlist
\item
  The genetic variant is exhibiting a vertical pleiotropic effect on
  trait A and the expression level of gene X. Here, all other
  assumptions being met, the genetic variant would be a valid instrument
  if used to estimate the causal effect of the gene expression level on
  trait A. However,
\item
  the SNP is exhibiting a horizontal pleiotropic effect on trait A and
  trait B through the two gene expression levels. Here the genetic
  variant would be invalid if used to estimate the causal effect of
  trait A on trait B.
\end{itemize}

\includegraphics{images/pleiotropy.png}

Adapted from Figure 3 in Hu et al (2016).

@@

The use of summary data in two-sample MR is described in detail
elsewhere (12,27). A brief outline of the procedure is as follows.
First, genetic instruments for the exposure trait need to be identified
- those SNPs with \(p < 10^{-8}\) are retained in order to ensure that
the first assumption of MR (that the instrument associates with the
exposure) is generally satisfied. We collect their effect sizes,
standard errors and effect alleles for the association with the exposure
trait. These can be obtained from published GWAS summary data, either
from from curated lists of established GWAS associations (10) or study
websites comprising complete summary data (all SNP effects across the
genome regardless of strength of association). Next, the effects of
those SNPs on the outcome need to be obtained, typically necessitating
access to complete summary data because these SNPs will not usually
reach genome-wide significance for the outcome trait (and are therefore
usually absent in curated catalogues).

@@

The Rucker framework (28), adapted to MR (6) uses estimates of
heterogeneity in the IVW and Egger frameworks to navigate between these
nested models. A jackknife approach (random selection with replacement
of the complete set of instruments) can be used to obtain a sampling
distribution for the model estimate amongst these four variations. Using
1000 rounds of jackknife estimates, we can obtain a final estimate using
the mean or the median of the distribution. We only use the jackknife
approach for MR analyses where there are 15 or more instruments, in
order to avoid saturating the possible number of instrument
combinations.

@@

\textbf{Median-based methods:} An alternative approach is to take the
median effect of all available instruments (5,29). This has the
advantage that only half the instruments need to be valid, and the
estimate will remain unbiased. The weighted median estimate develops the
approach further to allow stronger instruments to contribute more
towards the estimate can be obtained by weighting the contribution of
each instrument by the inverse of its variance. The penalised weighted
median estimator introduces a further weight to the instruments,
penalising any instrument that contributes substantially towards the
heterogeneity statistic. Together, this provides three median based
estimators. Other estimation strategies not considered here, such as
LASSO regression, have also been developed for the situation where at
least half of the instruments are valid (30).

@@

The simplest approach to selecting instruments is to take SNPs that have
been declared significant in the published GWAS for the exposure. This
typically involves obtaining SNPs that surpass \(p < 5 \times 10^{-8}\),
using clumping to obtain independent SNPs, and then replicating in an
independent sample. These results are often recorded in public GWAS
catalogs. Alternatively the clumping procedure can be performed using
complete summary data in MR-Base (12). Complete summary data refers to
association results for all SNPs used in a GWAS i.e.~not only those
passing statistical thresholds for significance. We call this the
``tophits'' strategy.

@@

because the \(cor(g_{i}, B)^2 = cor(A, B)^{2} cor(g_{i}, A)^{2}\). This
simple inequality will not hold in some cases, for example
\(\rho_{x, x_o} < \rho_{x,y}\rho_{y,y_o}\) where \(\rho_{x, x_o}\) and
\(\rho_{y, y_o}\) are the precision of the measurements of the \(x\) and
\(y\). Some combinations of confounding effects can also distort the
\(\rho_{g,x}\) and \(\rho_{g,y}\) parameters, as has been discussed in
detail previously (8). However, we use it here as a computationally
inexpensive and approximate method to identify variants that are likely
to be invalid. Steiger's Z-test of correlated correlations (31) can be
used to formally test the extent to which the two correlations are
statistically different.

Here we adapt this approach to automatically filter SNPs that are liable
to be invalid (Figure 1a). Other methods have been developed to identify
invalid instruments for the purposes of exclusion from MR analyses
(9,13,32), based on the notion that outlying instruments are more likely
to be a source of HP. A potential drawback of this approach is that the
outlier SNPs in a summary-set might be the only reliable ones. This
could arise, for example, for the \emph{CRP} variant instrumenting
C-reactive protein levels (21), or the \emph{SLC2A9} variant influencing
serum urate levels (33).

@@

We consider the 14 MR methods described above, for which instruments can
be supplied using two instrument selection strategies, leading to 28
methods in total (Supplementary table 2). Each method assumes or
performs best for a different model of pleiotropy. Our objective is to
select the method most likely to be correct for a specific MR analysis
by predicting the model of pleiotropy, and relating that prediction to
the most appropriate model. To achieve this we use a ``mixture of
experts'' machine learning approach (34), where each MR method is
considered to be an `expert', taking a summary-set as its input. An
overview of the approach is shown in Figure 2.

@@

It should be noted that in the original hierarchical mixture of experts
approach of (34) the gating function and the experts share the same
input space and are trained simultaneously using
expectation-maximisation. In our approach the gating function is defined
over a separate input space consisting of the parameter space of the
experts, and is trained separately in a supervised setting using
simulated data. As such our MR-MoE approach fits the framework of
meta-learning (35--37) where machine learning methods are applied at the
`meta-level' to analyse the results of machine learning experiments at
the `base-level', in order to be able to understand and model the
capabilities of each expert and recommend which expert should be applied
to a given problem. From this perspective the 28 MR methods are the
base-models, the gating function is the meta-model, and its
meta-features are the parameters of the summary-sets.

@@

We applied our analysis using summary data for 2407 phenotypes,
including 149 complex traits and diseases, 575 metabolites (38,39) and
1683 plasma protein levels (40). For the protein levels only GWAS
significant hits were available, so they could only be evaluated as
exposure phenotypes. The complex traits and diseases and metabolite
levels had complete summary data available obtained via MR-Base, and
thus they could be evaluated as both exposures (if they had significant
instruments) and outcomes. Together, we evaluated 715681 relationships.
The majority of these associations could only be evaluated using fewer
than 5 SNPs, and so the Wald ratio or IVW fixed effects methods were
used, but for 61029 associations the MR-MoE approach was applied.

@@

As an example of the database we performed a look up of associations
between all traits and LDL cholesterol (as measured in the GLGC
consortium (41)), where the MoE result had a false discovery rate of
0.05 using the MoE approach. This returned 287 putative associations,
amongst which 111 involved LDL cholesterol influencing other traits and
176 involved other traits influencing LDL cholesterol. A large
proportion of these traits were metabolites, which are dominated by
lipid fractions. Filtering to exclude the metabolomic studies ((38,39))
returned 27 associations with 23 traits (Figure 4a). There was a strong
association with coronary heart disease (0.41 log(OR) per SD, 0.31-0.52,
\(p=2.7\times10^{-12}\)). Here the IVW random effects method was chosen
after Steiger filtering, indicating the presence of HP being present
amongst the instruments.

We also performed a look up of associations that causally influence
years of schooling (42), or that years of schooling itself causally
influences. Using a false-discovery rate of 0.05, 45 traits were
returned as having some direction of causality with years of schooling
(Figure 4b). Several of these putative relationships appear to be
plausible, for example the influence of years of schooling on lower risk
of Alzheimer's disease, which has recently been reported (43). However,
it is clear that there are serious limitations in relying on MR as a
panacea for causal inference. For example, bi-directional causal
relationships with college completion exist because of the similarity in
trait definitions; and there are several traits (e.g.~childhood
intelligence and birth weight) which appear to be causally downstream of
years of schooling despite this being temporally impossible. Though
explanations could be conjured for this association, for example
parents' schooling influences childhood intelligence, or the genetic
instruments are shared across the two traits due to a third unmeasured
trait, an intuitive interpretation of causality cannot be applied
because the genetic instruments used proxy the liability to achieve some
level of educational attainment, and not necessarily the experience of
having attained a particular level of edication.

@@

Motivated by a recent flurry of activity in MR methodology development,
we have devised a machine learning approach, MR-MoE, that seeks to
predict the performance of each MR method, selecting the one which is
most likely to maximise power whilst minimising FDR for a specific
summary-set. The mixture of experts approach that we present here is
trained using random forest decision trees applied to extensive and
diverse simulations, and we demonstrate that it makes substantial
performance improvements over any other single method in the presence or
absence of extensive horizontal pleiotropy. As such, our method
contributes to the field of meta-learning but also to the rapidly
growing field of automatic model selection and hyper-parameter
optimization in machine learning (44). We applied the method to the
curated set of GWAS summary data present in MR-Base (12), generating
several million MR estimates. MR-MoE selected a method that indicated
some pattern of horizontal pleiotropy for almost 90\% of cases,
reinforcing the notion that it is the rule rather than the exception.

The trend of increasing GWAS sample sizes continues, but while the
opportunity that this affords MR to be furnished with more instruments
is typically welcomed, here we have shown that it is invalid
instruments, exhibiting horizontal pleiotropy or reverse causation, that
are more likely to be identified than valid ones. Strategising how MR is
to be used in practice, therefore, must consider horizontal pleiotropy
to be the rule rather than the exception. Instrumental variables in MR
are typically chosen blind, with GWAS significance being the only
criterion. We illustrated how confounders and reverse causal
associations can easily lead to invalid instrument selection in MR and
that most randomly generated patterns of horizontal pleiotropy cannot be
adequately accounted for by any single MR method.

Methods have been introduced recently that seek to improve MR estimates
through invalid instrument detection and removal (9,13,32). These attack
the problem from the perspective that variants which have a relatively
large influence on the causal estimate, or provide substantially
different estimates from the other variants, are likely to be exhibiting
HP, and their removal from the analysis may be warranted. A potential
drawback of this approach is that the outlier SNPs in a summary-set
might be the only reliable ones. This could arise, for example, for the
\emph{CRP} variant instrumenting C-reactive protein levels (21), because
CRP itself is downstream of many cellular functions and most GWAS
signals will not be specific for the trait. Steiger filtering has
attempted to sidestep this issue in removing invalid instruments,
seeking to detect those instruments which have associations with the
outcome that are too large to be consistent with the exposure causing
the outcome. We acknowledge that unmeasured confounding and measurement
error can introduce problems with Steiger filtering (8), but note that
our simulations indicate that all individual MR methods improve in
performance following its application.
